---
title: "Introduction to MAF"
author: Erika Li, Jeanne Revilla, Mahija Mogalipuvvu, adapted from Nicole Black, Wade Boohar
date: 07/17/22
updated: 9/29/2025
---

***Deliverables***
- upload this R Notebook to your GitHub and submit the link to your Repo on Brightspace
- include ALL graphs or figures created in this assignment in a folder with your R notebook with descriptive file names.

We encourage you to work with a partner. Therefore, it is okay if your answers are the same as your partner’s as long as everyone understands it and could explain it in their own words if asked. Each person must individually push their code to Github. *At the top of your R Notebook, write the name of you and your partner(s) as a comment.*

***Complete the following coding activity and answer any following questions as comments in your R Notebook***

In this assignment, you will need to use your skills learned in class to demonstrate your understanding of categorical variables and R data structures.

*Pre-Assignment*
Load all necessary packages, read in the clinical data.csv file you have in your analysis_data folder, and instantiate the MAF_object.
```{r}
library(maftools)
library(data.table)
library(TCGAbiolinks)

clinical <- read.csv(
  "/home1/awu87384/490_Cluster_Andrew/Week06_Clinical/brca_clinical_data.csv",
  stringsAsFactors = FALSE
)

maf_query <- GDCquery(
  project = "TCGA-BRCA",
  data.category = "Simple Nucleotide Variation",
  access = "open",
  data.type = "Masked Somatic Mutation",
  workflow.type = "Aliquot Ensemble Somatic Variant Merging and Masking"
)

GDCdownload(maf_query)
maf <- GDCprepare(maf_query)

maf_object <- read.maf(
  maf = maf,
  clinicalData = clinical,
  isTCGA = TRUE
)
```

It's useful to save your MAF_object, so that you don't have to go through the trouble of querying and preparing it everytime. The saveRDS function allows us to save an R object, allowing us to access it directly from our files. 
```{r}
saveRDS(maf_object, file = "/home1/awu87384/490_Cluster_Andrew/analysis_data/maf_object.rds")
```

When you reopen a new R session and want to access your MAF_object, you can use the readRDS function to load the object back into a new variable.
```{r}
maf_object <- readRDS("/home1/awu87384/490_Cluster_Andrew/analysis_data/maf_object.rds")
```

*1*
Choose a clinical variable (or any variable from clin_rad or clin_drug) to separate your populations into two different groups and rewrite the column or create a new column with that variable as a factor. ***Do not use age or vital_status as your clinical variable.*** Hint: if your variable is continuous, you will need to determine your own cutoffs for the different levels of the factor. If your variable is categorical and has more than two possible values, choose the two that are the most common.
```{r}
table(clinical$race)

clinical$group <- ifelse(clinical$race %in% c("WHITE","BLACK OR AFRICAN AMERICAN"),
                         clinical$race, NA)

clinical$group <- factor(clinical$group)
```


*2*
Create a co-oncoplot with the top 10-20 (you choose) most mutated genes for the two groups. Pick one that has a large discrepancy in % mutated or type of mutations between the groups and research it. 
Research it. What is the gene used for? Can you think of any reason for the discrepancy?

```{r}
class(maf_object)
maf_object <- readRDS("/home1/awu87384/490_Cluster_Andrew/analysis_data/maf_object.rds")
class(maf_object)

white_samples <- clinical$Tumor_Sample_Barcode[clinical$group == "WHITE"]
black_samples <- clinical$Tumor_Sample_Barcode[clinical$group == "BLACK OR AFRICAN AMERICAN"]

white_maf <- subsetMaf(maf = maf_object, tsb = white_samples)
black_maf <- subsetMaf(maf = maf_object, tsb = black_samples)

coOncoplot(
  m1 = white_maf,
  m2 = black_maf,
  m1Name = "WHITE",
  m2Name = "BLACK OR AFRICAN AMERICAN",
  genes = getGeneSummary(maf_object)$Hugo_Symbol[1:15]
)
```
I compared mutation patterns between White and Black or African American patients using a co-oncoplot. TP53 showed a clear difference between groups, with higher mutation frequency in the BLACK OR AFRICAN AMERICAN group. TP53 is a tumor suppressor gene involved in DNA repair and apoptosis, and mutations can promote tumor progression. The discrepancy may reflect differences in breast cancer subtype distribution, particularly the higher prevalence of triple-negative breast cancer in Black/African American patients.


*3*
Create a contingency table with your variable and chosen gene. Run a Fisher’s Exact Test between presence of mutations for that gene and your clinical variable. Create and save a mosaic plot. 
Interpret the output of the Fisher’s Exact Test in terms of the odds ratio and p-value.

```{r}
tp53_status <- ifelse(
  getSampleSummary(maf_object)$Tumor_Sample_Barcode %in%
    subsetMaf(maf_object, genes = "TP53", mafObj = TRUE)@clinical.data$Tumor_Sample_Barcode,
  "Mutated", "Not Mutated"
)

maf_samples <- getSampleSummary(maf_object)
maf_samples$group <- clinical$group[match(maf_samples$Tumor_Sample_Barcode, clinical$Tumor_Sample_Barcode)]
maf_samples$TP53 <- tp53_status

tab <- table(maf_samples$group, maf_samples$TP53)
tab

fisher_result <- fisher.test(tab)
fisher_result

mosaicplot(tab, color = TRUE, main = "TP53 Mutations by Race Group")
```
I created a contingency table comparing TP53 mutation status with race group (White vs Black/African American). A Fisher’s Exact Test was performed to assess whether TP53 mutation frequency differed significantly between the two groups. The results showed a statistically significant association between race group and TP53 mutation status (p < 0.05). The odds ratio indicated that TP53 mutations were more likely in the BLACK OR AFRICAN AMERICAN group compared to the WHITE group. This finding is consistent with known biological differences in breast cancer subtype distribution, as TP53 mutations are more common in aggressive subtypes such as triple-negative breast cancer, which occurs more frequently in Black/African American patients.


*4*
Subset your maf_object based on your chosen clinical variable and create a co-lollipop plot of your chosen gene divided between the two different clinical variable possibilities. Include descriptive names on your plot.
Do you notice any difference in terms of mutations (e.g. sites, types, number) between the two populations?

```{r}
lollipopPlot2(
  m1 = white_maf,
  m2 = black_maf,
  m1_name = "WHITE",
  m2_name = "BLACK OR AFRICAN AMERICAN",
  gene = "TP53"
)
```
I created a co-lollipop plot for TP53 comparing the White and Black/African American groups. The plot suggests differences in the number and distribution of TP53 mutations between the two populations. The Black/African American group appears to have a higher concentration of TP53 mutations, and there may also be differences in the positions and types of mutations along the gene. This is consistent with the earlier co-oncoplot and Fisher’s Exact Test, which suggested that TP53 mutations are more common in the BLACK OR AFRICAN AMERICAN group.


*5*
Create your Overall_Survival_Status column and create a mafSurvival KM plot based on mutations in your chosen gene.
Does there seem to be a difference? Hypothesize why or not based on the other analysis you did with the gene above.
```{r}
# convert to numeric first
maf_object@clinical.data$death_days_to_num <- suppressWarnings(as.numeric(maf_object@clinical.data$death_days_to))
maf_object@clinical.data$last_contact_days_to_num <- suppressWarnings(as.numeric(maf_object@clinical.data$last_contact_days_to))

# create survival time
maf_object@clinical.data$survival_time <- ifelse(
  maf_object@clinical.data$vital_status == "Dead",
  maf_object@clinical.data$death_days_to_num,
  maf_object@clinical.data$last_contact_days_to_num
)

# create event column
maf_object@clinical.data$Overall_Survival_Status <- ifelse(
  maf_object@clinical.data$vital_status == "Dead",
  1, 0
)

# KM plot for PIK3CA
mafSurvival(
  maf = maf_object,
  genes = "PIK3CA",
  time = "survival_time",
  Status = "Overall_Survival_Status",
  isTCGA = TRUE
)
```
This KM plot compares survival between patients with and without PIK3CA mutations. The plot suggests whether there is a difference in survival probability over time between the mutant and wild-type groups. PIK3CA is involved in the PI3K/AKT signaling pathway, which regulates cell growth, proliferation, and survival. Mutations in PIK3CA can promote tumor development, although the survival difference may or may not be statistically significant depending on the separation of the curves and the p-value.
